The force of mortality by life lived is the force of increment by life left in stationary populations

### 2015-02-09 LifetableIncrement/ReplyReviewers.tex

\documentclass{article}

\usepackage{amsmath}
\usepackage{natbib}
\bibpunct{(}{)}{,}{a}{}{;} 
\newcommand{\dd}{\; \mathrm{d}}
\usepackage{url}
\begin{document}
\title{Reply to Reviewers of submission 2518: The force of mortality by life
lived is the force of increment by life left in stationary populations}
\author{Author Redacted}

Thanks for the useful comments, which improved the paper. I
hope that concerns are now met with the current revision.

Reviewer A gave math and notation suggestions, which were correct and which I
heeded. I think the primary concern is mended by changing former equation (4)
(now equation (5)) to:
\begin{equation}
l^\star (y) = \int _\omega^y f(t) \dd t \notag \quad \text{.} 
\end{equation}

Reviewer B suggested I add detail to the applications section. Admittedly, these
applications have to do with the thanatological age perspective in general
rather than the neatness of symmetry with respect to chronological decrement and
thanatological renewal within the lifetable. I hope my suggested applications
are an acceptable ersatz. The specific case I suggest is for the measurement of
morbidity patterns in order to approach the question of whether or not there
has been compression of morbidity. This question has been asked many times, and
it appears that the usual approach is to simply compare $e(65)$ with average
disability rates at ages $65+$. That is a very rough gauge. Measuring morbidity
in thanatological age would help disentangle morbidity change from lifespan change,
which is what we really want in order to say something about compression.

Reviewer C likened the thanatological perspective change to viewing the contents
of a room from above. This is a good analogy in the case of stationary
populations, but note that it only works for the marginal profile of lifetable
functions. The perspective change is more complex than that, bearing in mind
that the functions are decomposable into chronological and thanatological age
(described for the case of the pyramid). For an image of such heterogeneity, see
the colored images in this abstract to PAA, 2014: \url{http://paa2014.princeton.edu/papers/141036}
(before I knew about \citep{brouard1986structure}). I don't think an image is
necessary in this paper, to appreciate this, though, but I will oblige if the
wish is echoed further. I added a line at the end of the proof section ``The
main contribution of this relationship is to point out the symmery of chronological decrement and thanatological renewal processes in
the case of stationary populations.'' This contribution isn't revolutionary or
of great use, but the lifetable, stationary and stable populations, and the
renewal model stand at the core of demography, which ought to be
solid. I think tiny observations about them are worthy of publishing, and moreso
in this series, which values aesthetics. The actual main contribution of this
article is just to make a formal demographer smile.

Note that the stationary age structure finding, on which this result builds, has
appeared in the literatute again, since this paper was submitted
\citep{rao2014generalization}. These and previous authors were apparently also
not aware of Brouard's contributions. I hope to subtly set the record straight.

\bibliographystyle{plainnat}
  \bibliography{bibliography}

\end{document}

### 2015-02-09 LifetableIncrement/LifetableIncrement.tex
added rao carey, Fries. Also reply to reviewers. main submission modified

%\chead{submitted as a relationship report to be included in the Special
%Collection ``Formal Relationships'' edited by Joshua R. Goldstein and James W.
%Vaupel}
%\lfoot{Submission \#2518 for review}
%\rfoot{}
\fancyhf{}
\fancyhead[C]{\footnotesize{submitted as a relationship report to be included in the
Special Collection ``Formal Relationships'' edited by Joshua R. Goldstein and James W.
Vaupel}}
\fancyfoot[R]{\footnotesize{\bfseries{page \thepage}}}
\fancyfoot[L]{\footnotesize\bfseries{Submission \#2518 for review}}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrulewidth}{0pt}
%\setlength\headheight{117.89105pt}
\setcounter{page}{2}
\pagestyle{fancy}

@@

Age can be defined as chronological (time since birth) or thanatological (time
until death). In a stationary population, decrement over
chronological age, as described by the lifetable, is
identical to increment over thanatological age.

@@

Chronological age structure equals thanatological age structure in a stationary
population \citep{brouard1989mouvements,vaupel2009life}:

\begin{equation}
\label{eq:vaupel}
c(a) = g(a) \quad\quad \text{,}
\end{equation}
where $c(a)$ is the stable chronological age structure expressed as a
proportion, and $g(a)$ is proportion of the population with $a$ remaining years
of life. Using $a$ to index chronological age and $y$ to index
thanatological age, the amount of time left until death. Equation
\eqref{eq:vaupel} is the same as:

\begin{equation}
\label{eq:equivalent}
c(a) = c^\star (y) \quad \text{for}\quad a = y \quad\quad\text{,}
\end{equation}
where $c^\star (y)$ refers to the stable thanatological age
structure. $c(a)$ is proportional to the survival
function, $l(a)$, which means that $l(a)$ is proportional to some
thanatological function, $l^\star (y)$, a stationary thanatological cumulative
\textit{increment} function. To untangle how this is so, it helps to be
explicit about what $l(a)$ represents. A birth cohort is a
group of individuals that are born at the same point in time and experience attrition over age until extinction, as described by $l(a)$. $l(a)$ is also
the sum of all deaths to the birth cohort at ages greater than $a$, and can be
thought of as the de-accumulation of the stock of future deaths over age.

\begin{equation}
l(a) = \int _{0}^\infty f(a+y) \dd y \quad \text{,}
\end{equation}

\noindent where $f(a)$ is the lifetable density function, often denoted by
$d(a)$. $f(a)$ gives the probability that a member of the birth cohort born in year $t$
will die in the year $t+a$. The deaths that will occur together in year
$t+a$ comprise a \textit{death cohort}. In the next year $t+1$,
$l(0)\cdot f(a-1)$ births increment to the $t+a$ death cohort in the stationary
population. Death cohorts grow monotonically in this way, starting with a few
members that will enjoy the maximum attainable lifespan, $ y = \omega$, and
accumulating new members as time approaches $t_0+\omega$ and $y$ decreases
toward 0:

\begin{equation}
l^\star (y) = \int _\omega^0 f(y) \dd y \quad \text{.}
\end{equation}

\noindent Extinction of the death cohort is simultaneous upon reaching
thanatological age 0. The lifetable deaths distribution is under this
perspective a distribution of the births to a death cohort, when read
from the highest to the lowest ages. The rate at which new births accumulate to
death cohorts over thanatological age, $\mu^\star (y)$ is given by
\begin{align}
\mu ^\star (y) &= \frac{-l^{\star^\prime} (y)}{l^\star (y)} \quad \text{,}
               \intertext{and since $l(a) = l^\star{a}$:}
               &= \frac{-l^\prime (y)}{l(y)} \notag\\
               &= \mu (y)
\end{align}

The rate of birth accumulation into death cohorts over thanatological age is
equal to the rate of death attrition of birth cohorts over chronological age in
stationary populations. The stock of deaths in the future for
chronological age $a$, $l(a)$, is symmetrically a description of births in the
past when structured by thanatological age, $l^\star(y)$. The remaining
lifetable columns are subject to similar reinterpretations when viewed under
thanatological age.

@@

The term ``thanatological age'' has not previously appeared in the
literature, and was coined by Ken Wachter sometime prior to 2001. Thanatos was
the Greek god of death, which is used as the reference point for age under this
perspective. Explicit decompositions of chronological
age groups into remaining lifespan classes is to our knowledge only found in
\citet{brouard1986structure}, who redistributed
population pyramids by remaining years of life, and
\citet{miller2001increasing}, who examined medicare expenditure as a function of time until death using the
same lifetable identities. Equivalence of chronological and
thanatological age structures in stationarity, also known as Carey's equality,
is proven by \citet{brouard1989mouvements} and again in
 \citet{vaupel2009life}, which added detail to the more summary result reported
 in \citet{goldstein2009life}, that mean (or total) remaining lifespan is equal to mean (or total) life lived in the stationary population.
These latter two papers and \citet{goldstein2012historical} describe more of the
lineage of result presented here.

@@

\section*{Applications}
Thanatological age structure can be applied to stable populations (subject to a
growth rate, $r$), though we leave the description of a thanatological
renewal model for future work. Thanatological age equates individuals that share
a common terminal state rather than a common origin state. Here this is
the absorbing state of death, but the method generalizes to any terminal state
or lifecourse transition that can be modeled using lifetable techniques. Potential area
applications that may gain insights using such remaining-time methods
include morbidity, disability, late-life savings and investment behavior, or perhaps time to birth, menopause, retirement, or graduation.

>>>>

%\chead{submitted as a relationship report to be included in the Special
%Collection ``Formal Relationships'' edited by Joshua R. Goldstein and James W.
%Vaupel}
%\lfoot{Submission \#2518 for review}
%\rfoot{}
\fancyhf{}
\fancyhead[C]{\footnotesize{submitted as a relationship report to be included in the
Special Collection ``Formal Relationships'' edited by Joshua R. Goldstein and James W.
Vaupel}}
\fancyfoot[R]{\footnotesize{\bfseries{page \thepage}}}
\fancyfoot[L]{\footnotesize\bfseries{Submission \#2518, revised}}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrulewidth}{0pt}
%\setlength\headheight{117.89105pt}
\setcounter{page}{2}
\pagestyle{fancy}

@@

Age can be defined as chronological, $a$ (time since birth), or thanatological,
$y$ (time until death). In a stationary population, decrement over
chronological age, as described by the lifetable, is
identical to increment over thanatological age. Define the chronological force
of mortality, $\mu(a)$, and the thanatological force of increment,
$\mu^\star(y)$. We have:
\begin{equation}
\mu(a) = \mu^\star(y)\quad\quad\text{for~}a=y\quad \text{.}
\end{equation}

@@

Chronological age structure equals thanatological age structure in a stationary
population \citep{brouard1989mouvements,vaupel2009life,rao2014generalization}:

\begin{equation}
\label{eq:vaupel}
c(a) = g(a) \quad\quad \text{,}
\end{equation}
where $c(a)$ is the stable chronological age structure expressed as a
proportion, and $g(a)$ is the proportion of the population with $a$ remaining
years of life. Using $a$ to index chronological age and $y$ to index
thanatological age (time left until death), equation \eqref{eq:vaupel} is the
same as:

\begin{equation}
\label{eq:equivalent}
c(a) = c^\star (y) \quad \text{for}\quad a = y \quad\quad\text{,}
\end{equation}
where $c^\star (y)$ refers to the stable thanatological age
structure. $c(a)$ is proportional to the survival
function, $l(a)$, which means that $c^\star(y)$ is proportional to some
thanatological function, $l^\star (y)$, a stationary thanatological cumulative
\textit{increment} function. To untangle how this is so, it helps to be
explicit about what $l(a)$ represents. A birth cohort is a
group of individuals that are born at the same point in time and experience attrition over age until extinction, as described by $l(a)$. $l(a)$ is also
the sum of all deaths to the birth cohort at ages greater than $a$, and can be
thought of as the de-accumulation of the stock of future deaths over age.

\begin{equation}
l(a) = \int _{0}^\omega f(a+t) \dd t \quad \text{,}
\end{equation}

\noindent where $f(a)$ is the lifetable density function, often denoted by
$d(a)$. $f(a)$ gives the probability that a member of the birth cohort born in year $t$
will die in the year $t+a$. The deaths that will occur together in year
$t+a$ comprise a \textit{death cohort}. In the next year $t+1$,
$l(0)\cdot f(a-1)$ births increment to the $t+a$ death cohort in the stationary
population. Death cohorts grow monotonically in this way, starting with a few
members that will enjoy the maximum attainable lifespan, $ y = \omega$, and
accumulating new members as time approaches $t+\omega$ and $y$ decreases
toward 0:

\begin{equation}
l^\star (y) = \int _\omega^y f(t) \dd t \quad \text{.}
\end{equation}

\noindent Extinction of the death cohort is simultaneous upon reaching
thanatological age 0. The lifetable deaths distribution is under this
perspective a distribution of the births to a death cohort, when read
from the highest to the lowest ages. The rate at which new births accumulate to
death cohorts over thanatological age, $\mu^\star (y)$ is given by
\begin{align}
\mu ^\star (y) &= \frac{-l^{\star^\prime} (y)}{l^\star (y)} \quad \text{,}
               \intertext{and since $l(a) = l^\star(a)$:}
               &= \frac{-l^\prime (y)}{l(y)} \notag\\
               &= \mu (y)
\end{align}

The rate of birth accumulation into death cohorts over thanatological age is
equal to the rate of death attrition of birth cohorts over chronological age in
stationary populations. The stock of deaths in the future for
chronological age $a$, $l(a)$, is symmetrically a description of births in the
past when structured by thanatological age, $l^\star(y)$. The remaining
lifetable columns are subject to similar reinterpretations when viewed under
thanatological age. The main contribution of this relationship is to point out
the symmery of chronological decrement and thanatological renewal processes in
the case of stationary populations.

@@

The term ``thanatological age'' has not previously appeared in the
literature, and was coined by Ken Wachter sometime prior to 2001. Thanatos was
the Greek god of death, which is used as the reference point for age under this
perspective. Explicit decompositions of chronological
age groups into remaining lifespan classes is to our knowledge only found in
\citet{brouard1986structure}, who redistributed
population pyramids by remaining years of life, and
\citet{miller2001increasing}, who examined medicare expenditure as a function of time until death using the
same lifetable identities. Equivalence of chronological and
thanatological age structures in stationarity, also known as Carey's equality,
is proven by \citet{brouard1989mouvements}, again in
 \citet{vaupel2009life},\footnote{\citet{vaupel2009life} added detail to the
 more summary result reported in \citet{goldstein2009life}, which showed that
 the mean (or total) remaining lifespan is equal to mean (or total) life lived in the stationary population.} and in more detail by \citet{rao2014generalization}.
These latter two papers and \citet{goldstein2012historical} describe more of the
lineage of the result presented here.

@@

The thanatological age perspective only offers the kind of profile symmetry
presented in this paper for the theoretical case of stationary populations. For
changing populations, the chronological and thanatological age perspectives
typically offer different profiles of the same phenomena, due to changes in
lifespan distributions and fluctuations in the birth flow, and therefore offer
complementary information on population structure.

\section*{Applications}
Thanatological age structure can be applied to stable populations (subject to a
growth rate, $r$), though we leave the description of a thanatological
renewal model for future work. Thanatological age equates individuals that share
a common terminal state rather than a common origin state. In the present
relationship, this is the absorbing state of death, but the method generalizes to any terminal state
or lifecourse transition that can be modeled using lifetable techniques. Potential area
applications that may gain insights using such remaining-time methods
include morbidity, disability, late-life savings and investment behavior, or
perhaps time to birth, menopause, retirement, or graduation. Glacial or open
ice pack, old growth forest, and prison populations are other examples of
aggregates for which remaining time structure is inherently of equal or greater
interest than time passed. Populations of fixed or controlled size, or where
entries are largely a function of exits are also prime candidates for analysis under some
analogy to thanatological age. Examples of such populations include professional
athletes in leagues, tenured professors, company directors, vehicle fleets.

As a specific example, the question of morbidity compression has often been
posed as a matter of comparing age of onset with remaining life expectancy
\citep[e.g., ][]{fries2002aging,fries2003measuring}.\footnote{In the health
literature, one often sees $e(65)$ compared with average disability levels above age 65.} In
retrospect (within death cohorts), one can analytically separate between changes
in lifespan and changes in morbidity as a function of time until death. Cross-tabulations of morbidity by chronological and thanatological age for two birth cohorts would allow the
researcher to estimate changes in the morbidity profile over time free from
distortion due to changes in the lifespan distribution, and so directly settle
the question of morbidity compression.
